---
title: "Bike Collision GIS analysis"
author: "Ravi Brenner"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(sf)
library(ggspatial)
library(sfhotspot)
```

Importing and cleaning data

Bike routes
```{r}
bike_routes <- st_read("data/nyc_bike_routes.shp")  |>
  st_transform(crs = 2263) |>
  filter(year(instdate) <= 2024,
         status == "Current") |> # only want bike routes that existed in 2024 or earlier
  mutate(ft_facilit = case_match(ft_facilit,
                                 "<Null>" ~ NA,
                                 "NA" ~ NA,
                                 .default = ft_facilit),
         tf_facilit = case_match(tf_facilit,
                                 "<Null>" ~ NA,
                                 "NA" ~ NA,
                                 .default = tf_facilit),
         boro = case_match(boro,
                           1 ~ "Manhattan",
                           2 ~ "Bronx",
                           3 ~ "Brooklyn",
                           4 ~ "Queens",
                           5 ~ "Staten Island"),
         facilitycl = case_match(facilitycl,
                                 "I" ~ "Protected",
                                 "II" ~ "Conventional",
                                 "III" ~ "Shared Lane",
                                 "L" ~ "Link"),
         facilitycl = factor(facilitycl,
                             levels = c("Protected",
                                        "Conventional",
                                        "Shared Lane",
                                        "Link")))
```

Collisions for 2024
```{r}
collisions_2024 <- read_csv("data/collisions_2024.csv",
                            na = c("NA")) |>
  drop_na(latitude,longitude) |>
  filter(latitude != 0,
         longitude != 0,
         number_of_cyclist_injured > 0 | number_of_cyclist_killed > 0) |>
  st_as_sf(coords = c("longitude","latitude"),
           crs = 4326) |>
  st_transform(crs = st_crs(bike_routes))
```

Bike counts and counters
```{r}
bike_counts_2024 <- read_csv("data/bike_counts_2024.csv")

bike_counters <- read_csv("data/bike_counters.csv") |>
  filter(latitude != 0,
         longitude != 0) |>
  st_as_sf(coords = c("longitude","latitude"),
           crs = 4326) |>
  st_transform(crs = st_crs(bike_routes))
```

NYC boundaries
```{r}
nyc <- st_read("data/Counties_Shoreline.shp") |>
  filter(NAME %in% c("Bronx","Richmond","Kings","Queens","New York")) |>
  st_transform(crs = st_crs(bike_routes))
```

Community districts
```{r}
# populations
cd_pop <- readxl::read_excel("data/2020CensusData/nyc_decennialcensusdata_2010_2020_change-core-geographies_v2.xlsx",
                             sheet = "2020") |>
  janitor::clean_names() |> 
  filter(geo_type == "CD") |>
  mutate(geo_id = as.numeric(geo_id)) |>
  select(geo_id, pop1)

nyc_cd <- st_read("data/nyc_cd.shp") |>
  st_transform(crs = st_crs(bike_routes)) |>
  inner_join(cd_pop, 
             by = join_by(boro_cd == geo_id))
```



# Preliminary map

```{r}
ggplot() + 
  annotation_map_tile(type = "cartolight", zoomin = 0) +
  geom_sf(data = collisions_2024,
          color = "lightblue",
          alpha = 0.5) +
  geom_sf(data = bike_routes,
          aes(color = facilitycl)) +
  scale_color_discrete(name = "Bike Lane Type") + 
  labs(title = "Bike Lanes and Collisions Involving a Bicycle in New York City, 2024",
       caption = "Author: Ravi Brenner\nSource: NYC Open Data\nNote: crash locations shown in blue may represent multiple crashes at that location") + 
  theme_void()  +
  theme( 
    plot.caption.position = "plot",
    plot.caption = element_text(hjust = 1,
                                vjust = 20),
    legend.position = "inside",
    legend.position.inside = c(0.2,0.7),
    legend.key.height = unit(0.6, "cm"), 
    legend.text = element_text(size = 10),
    legend.background = element_rect(
      fill = alpha("white", 0.8),
      color = NA)
  ) + 
  annotation_north_arrow(
    height = unit(.8, "cm"),
    width = unit(.6, "cm"),
    pad_x = unit(2.5, "in"), 
    pad_y = unit(0.35, "in"),
    style = north_arrow_minimal
  ) +
  annotation_scale(
    style = "ticks",
    pad_x = unit(1.5, "in"),
    pad_y = unit(0.4, "in"),
    unit_category = "imperial",
    width_hint = 0.1
  )

ggsave(
  filename = "output/prelim_map_1.jpg",
  width = 7,
  height = 7
  )
```

join collisions with city neighborhoods
```{r}
collisions_cd <- st_join(nyc_cd, collisions_2024)

temp <- collisions_cd |>
  group_by(boro_cd) |>
  summarize(num_crashes = n()) |>
  mutate(area = as.numeric(st_area(geometry) / (5280*5280)),
         crashes_per_area = num_crashes / area) 

jenks_breaks <- classInt::classIntervals(temp$crashes_per_area, style = "jenks", n = 4)$brks

temp |>
  ggplot() + 
  annotation_map_tile(type = "cartolight", zoomin = 0) + 
  geom_sf(aes(fill = crashes_per_area),
          alpha = 0.5) + 
  scale_fill_viridis_c(name = "Crashes per Square Mile",
                       breaks = jenks_breaks,
                       labels = c(0.15,14.5,37.4,58.1,123.5)) +
  labs(title = "Bike Crashes by Community District, New York City, 2024",
       caption = "Author: Ravi Brenner\nSource: NYC Open Data") + 
  theme_void() +
    theme( 
    plot.caption.position = "plot",
    plot.caption = element_text(hjust = 1,
                                vjust = 15),
    legend.position = "inside",
    legend.position.inside = c(0.2,0.7),
    legend.key.height = unit(0.6, "cm"), 
    legend.text = element_text(size = 10),
    legend.background = element_rect(
      fill = alpha("white", 0.5),
      color = NA),
    legend.margin = margin(b = 5)
  ) + 
  annotation_north_arrow(
    height = unit(.8, "cm"),
    width = unit(.6, "cm"),
    pad_x = unit(2.5, "in"), 
    pad_y = unit(0.35, "in"),
    style = north_arrow_minimal
  ) +
  annotation_scale(
    style = "ticks",
    pad_x = unit(1.5, "in"),
    pad_y = unit(0.4, "in"),
    unit_category = "imperial",
    width_hint = 0.1
  )

ggsave(
  filename = "output/prelim_map_2.jpg",
  width = 7,
  height = 7
  )
```

# EDA maps
locations of bike counters
```{r}
ggplot() +
  annotation_map_tile(type = "cartolight", zoomin = 0) + 
  geom_sf(data = bike_counters) + 
  geom_sf(data = bike_routes,
          aes(color = facilitycl)) +
  theme_void()
```

Bike counters with counts
```{r}
bike_counts_2024 |> 
  summarize(.by = id, 
            counts = sum(counts) / 365) |>
  left_join(bike_counters, by = "id") |>
  st_as_sf() |>
  ggplot() + 
  annotation_map_tile(type = "cartolight", zoomin = 0) + 
  geom_sf(aes(color = counts)) +
  theme_void()
```

spatially join bike counters and lanes
```{r}
bike_counts_2024 |> 
          summarize(.by = id, 
                    counts = sum(counts) / 365) |>
          left_join(bike_counters, by = "id") |>
  st_as_sf() |>
  st_join(bike_routes |>
              st_buffer(dist = 50),
          left = TRUE) |> 
  ggplot() + 
  annotation_map_tile(type = "cartolight",zoomin = 0) + 
  geom_sf(aes(color = facilitycl)) + 
  theme_void()
```

So counters are (mostly) near protected bike lanes


spatially join crashes and bike lanes?
First need to add a buffer around lanes
```{r}
buffers <- st_buffer(bike_routes, dist = 100)

collisions_routes <- st_join(collisions_2024, buffers) |>
  mutate(near_bike_route = if_else(!is.na(segmentid),1,0)) |>
  filter(near_bike_route == 1)
```

```{r}
ggplot() +
  annotation_map_tile(type = "cartolight", zoomin = 0) + 
  geom_sf(data = collisions_routes,
          aes(color = facilitycl),
          alpha = 0.5) +
  geom_sf(data = bike_routes,
          aes(color = facilitycl)) +
  theme_void() 
```



Collisions in my CD
```{r}
collisions_cd <- st_join(collisions_2024,nyc_cd) 

bike_routes_cd <- st_join(bike_routes,
                          nyc_cd |>
                            filter(boro_cd == 107),
                          left = FALSE)

collisions_cd |>
  filter(boro_cd == 107) |>
  ggplot() + 
  annotation_map_tile(type = "cartolight", zoomin = 0) + 
  geom_sf() + 
  geom_sf(data = bike_routes_cd,
          aes(color = facilitycl)) + 
  theme_void()
```


Density of bike lanes by neighborhood
```{r}
bike_routes |>
  mutate(bike_lane_length = st_length(bike_routes)) |>
  group_by(boro) |>
  summarize(bike_lane_length = sum(bike_lane_length)) 

nyc |>
  mutate(area = st_area(nyc)) |>
  select(NAME, area)
```

## analysis ideas
- time period: 2024
- unit of analysis:
  - crash point. was crash in bike lane? what type of bike lane?
  - bike lane. how many crashes per mile of bike lane of different types? can we focus on high volume bike lanes?
  - neighborhood / CD. density of crashes and bike lanes per mile? as a fraction of total in the city?
  
I am worried about not being able to account for traffic volumes. of course there will be more crashes where there are more riders...Maybe focusing in on bike counters or a single neighborhood will help, but it is tricky.

# Main analysis

## Crashes by bike lane type

Bike crashes by lane type (raw)

Categorizing each crash by whether it was within 100 feet of a bike lane
```{r}
buffers <- st_buffer(bike_routes, dist = 100)

collisions_routes <- st_join(collisions_2024, buffers) |>
  mutate(near_bike_route = if_else(!is.na(segmentid),1,0)) 

collisions_routes |>
  group_by(facilitycl) |>
  summarize(n = n(),
            num_bike_injured = sum(number_of_cyclist_injured),
            num_bike_killed = sum(number_of_cyclist_killed)) |>
  st_drop_geometry() |>
  inner_join(bike_routes |>
               mutate(length = as.numeric(st_length(geometry))) |>
               group_by(facilitycl) |>
               summarize(length_miles = sum(length)/5280) |>
               st_drop_geometry()
             ) |>
  mutate(crashes_per_mile = n / length_miles,
         injuries_per_mile = num_bike_injured / length_miles,
         deaths_per_mile = num_bike_killed / length_miles) 
```

## crashes by bike lane - counters

focus on lanes near counters
```{r}
buffers <- st_buffer(bike_routes,100)

counters_routes <- st_join(buffers,bike_counters) |>
  filter(!is.na(counter))

counters_routes |>
  group_by(bikeid,id,counter) |>
  count() |>
  st_drop_geometry() |>
  left_join(
    bike_counts_2024 |>
      group_by(id) |>
      summarize(counts = sum(counts))
  ) |>
  distinct(bikeid,counter,counts) |>
  group_by(bikeid) |>
  summarize(total_count = sum(counts)) |>
  inner_join(collisions_routes |>
               st_drop_geometry() |>
               group_by(bikeid) |>
               summarize(n = n(),
                         num_bike_injured = sum(number_of_cyclist_injured),
                         num_bike_killed = sum(number_of_cyclist_killed))) |>
  mutate(crashes_per_count = n / total_count) |>
  inner_join(bike_routes) |> 
  st_as_sf() |>
  ggplot() +
  annotation_map_tile(type = "cartolight", zoomin = 0) +
  geom_sf(aes(color = 10000*crashes_per_count)) +
  scale_color_viridis_c(name = "Crashes per 10,000 riders") +
  theme_void()



```

## By neighborhood 

crashes and bike lanes per area (and per population?)

```{r}
cd_collisions <- nyc_cd |>
  filter(pop1 > 1000) |>
  mutate(area = as.numeric(st_area(geometry)) / (5280*5280)) |>
  st_join(bike_routes |>
  mutate(bike_lane_length = st_length(bike_routes))) |>
  group_by(boro_cd, area, pop1) |> 
  summarize(bike_lane_length = as.numeric(sum(bike_lane_length))) |>
  st_join(collisions_2024) |> 
  group_by(boro_cd, area, pop1, bike_lane_length) |>
  summarize(num_crashes = n()) |>
  mutate(crashes_per_mile = num_crashes/bike_lane_length,
         crashes_per_area = num_crashes / area,
         bike_lanes_per_area = bike_lane_length / area,
         crashes_per_pop = num_crashes / pop1,
         bike_lanes_per_pop = bike_lane_length / pop1) |>
  ungroup() |>
  mutate(crashes_pct = num_crashes / sum(num_crashes),
         bike_lanes_pct = bike_lane_length / sum(bike_lane_length),
         pop_pct = pop1 / sum(pop1)) 

cd_collisions |>
  ggplot() +
  annotation_map_tile(type = "cartolight", zoomin = 0) +
  geom_sf(aes(fill = crashes_per_area),
          alpha = 0.5) + 
  scale_fill_viridis_c() +
  theme_void()
```


Here, can see that as the bike lane density increases, the number of crashes increases.
this would probably indicate that bike lane presence itself leads to more riders, which then means higher crashes. Expected but important to note
```{r}
cd_collisions |>
  ggplot(aes(x = bike_lanes_per_area, y = crashes_per_area)) +
  geom_point() +
  geom_smooth(method = "lm")

summary(lm(crashes_per_area ~ bike_lanes_per_area, data = cd_collisions))
```

I think the outline should be something like this
- crashes by bake lane type (table)
- crahes per mile of bike lane at CD level (scatterplot)
- bike lanes per CD area
- bike lanes per CD pop
- crashes per CD area
- crashes per CD pop


