---
title: "Bike Collision GIS analysis"
author: "Ravi Brenner"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(sf)
library(ggspatial)
library(patchwork)
library(sfhotspot)
library(sfdep)
library(biscale)

```

Importing and cleaning data

Bike routes
```{r}
bike_routes <- st_read("data/nyc_bike_routes.shp")  |>
  st_transform(crs = 2263) |>
  filter(year(instdate) <= 2024,
         status == "Current") |> # only want bike routes that existed in 2024 or earlier
  mutate(ft_facilit = case_match(ft_facilit,
                                 "<Null>" ~ NA,
                                 "NA" ~ NA,
                                 .default = ft_facilit),
         tf_facilit = case_match(tf_facilit,
                                 "<Null>" ~ NA,
                                 "NA" ~ NA,
                                 .default = tf_facilit),
         boro = case_match(boro,
                           1 ~ "Manhattan",
                           2 ~ "Bronx",
                           3 ~ "Brooklyn",
                           4 ~ "Queens",
                           5 ~ "Staten Island"),
         facilitycl = case_match(facilitycl,
                                 "I" ~ "Protected",
                                 "II" ~ "Conventional",
                                 "III" ~ "Shared Lane",
                                 "L" ~ "Link"),
         facilitycl = factor(facilitycl,
                             levels = c("Protected",
                                        "Conventional",
                                        "Shared Lane",
                                        "Link")))
```

Collisions for 2024
```{r}
collisions_2024 <- read_csv("data/collisions_2024.csv",
                            na = c("NA")) |>
  drop_na(latitude,longitude) |>
  filter(latitude != 0,
         longitude != 0,
         number_of_cyclist_injured > 0 | number_of_cyclist_killed > 0) |>
  st_as_sf(coords = c("longitude","latitude"),
           crs = 4326) |>
  st_transform(crs = st_crs(bike_routes))
```

Bike counts and counters
```{r}
bike_counts_2024 <- read_csv("data/bike_counts_2024.csv")

bike_counters <- read_csv("data/bike_counters.csv") |>
  filter(latitude != 0,
         longitude != 0) |>
  st_as_sf(coords = c("longitude","latitude"),
           crs = 4326) |>
  st_transform(crs = st_crs(bike_routes))
```

NYC boundaries
```{r}
nyc <- st_read("data/Counties_Shoreline.shp") |>
  filter(NAME %in% c("Bronx","Richmond","Kings","Queens","New York")) |>
  st_transform(crs = st_crs(bike_routes))
```

Community districts
```{r}
cd_pop <- readxl::read_excel("data/2020CensusData/nyc_decennialcensusdata_2010_2020_change-core-geographies_v2.xlsx",
                             sheet = "2020") |>
  janitor::clean_names() |> 
  filter(geo_type == "CD") |>
  mutate(geo_id = as.numeric(geo_id)) |>
  select(geo_id, pop1)

nyc_cd <- st_read("data/nyc_cd.shp") |>
  st_transform(crs = st_crs(bike_routes)) |>
  inner_join(cd_pop, 
             by = join_by(boro_cd == geo_id))
```


ZIPs
```{r}
nyc_zip <- st_read("data/nyc_zips.shp") |>
  st_transform(crs = st_crs(bike_routes)) |>
  filter(zip != 99999)
```



# Preliminary map

```{r}
ggplot() + 
  annotation_map_tile(type = "cartolight", zoomin = 0) +
  geom_sf(data = collisions_2024,
          color = "lightblue",
          alpha = 0.5) +
  geom_sf(data = bike_routes,
          aes(color = facilitycl)) +
  scale_color_discrete(name = "Bike Lane Type") + 
  labs(title = "Bike Lanes and Collisions Involving a Bicycle in New York City, 2024",
       caption = "Author: Ravi Brenner\nSource: NYC Open Data\nNote: crash locations shown in blue may represent multiple crashes at that location") + 
  theme_void()  +
  theme( 
    plot.caption.position = "plot",
    plot.caption = element_text(hjust = 1,
                                vjust = 20),
    legend.position = "inside",
    legend.position.inside = c(0.2,0.7),
    legend.key.height = unit(0.6, "cm"), 
    legend.text = element_text(size = 10),
    legend.background = element_rect(
      fill = alpha("white", 0.8),
      color = NA)
  ) + 
  annotation_north_arrow(
    height = unit(.8, "cm"),
    width = unit(.6, "cm"),
    pad_x = unit(2.5, "in"), 
    pad_y = unit(0.35, "in"),
    style = north_arrow_minimal
  ) +
  annotation_scale(
    style = "ticks",
    pad_x = unit(1.5, "in"),
    pad_y = unit(0.4, "in"),
    unit_category = "imperial",
    width_hint = 0.1
  )

ggsave(
  filename = "output/prelim_map_1.jpg",
  width = 7,
  height = 7
  )
```

join collisions with city neighborhoods
```{r}
collisions_cd <- st_join(nyc_cd, collisions_2024)

temp <- collisions_cd |>
  group_by(boro_cd) |>
  summarize(num_crashes = n()) |>
  mutate(area = as.numeric(st_area(geometry) / (5280*5280)),
         crashes_per_area = num_crashes / area) 

jenks_breaks <- classInt::classIntervals(temp$crashes_per_area, style = "jenks", n = 4)$brks

temp |>
  ggplot() + 
  annotation_map_tile(type = "cartolight", zoomin = 0) + 
  geom_sf(aes(fill = crashes_per_area),
          alpha = 0.5) + 
  scale_fill_viridis_c(name = "Crashes per Square Mile",
                       breaks = jenks_breaks,
                       labels = c(0.15,14.5,37.4,58.1,123.5)) +
  labs(title = "Bike Crashes by Community District, New York City, 2024",
       caption = "Author: Ravi Brenner\nSource: NYC Open Data") + 
  theme_void() +
    theme( 
    plot.caption.position = "plot",
    plot.caption = element_text(hjust = 1,
                                vjust = 15),
    legend.position = "inside",
    legend.position.inside = c(0.2,0.7),
    legend.key.height = unit(0.6, "cm"), 
    legend.text = element_text(size = 10),
    legend.background = element_rect(
      fill = alpha("white", 0.5),
      color = NA),
    legend.margin = margin(b = 5)
  ) + 
  annotation_north_arrow(
    height = unit(.8, "cm"),
    width = unit(.6, "cm"),
    pad_x = unit(2.5, "in"), 
    pad_y = unit(0.35, "in"),
    style = north_arrow_minimal
  ) +
  annotation_scale(
    style = "ticks",
    pad_x = unit(1.5, "in"),
    pad_y = unit(0.4, "in"),
    unit_category = "imperial",
    width_hint = 0.1
  )

ggsave(
  filename = "output/prelim_map_2.jpg",
  width = 7,
  height = 7
  )
```

# EDA maps
locations of bike counters
```{r}
ggplot() +
  annotation_map_tile(type = "cartolight", zoomin = 0) + 
  geom_sf(data = bike_counters) + 
  geom_sf(data = bike_routes,
          aes(color = facilitycl)) +
  theme_void()
```

Bike counters with counts
```{r}
bike_counts_2024 |> 
  summarize(.by = id, 
            counts = sum(counts) / 365) |>
  left_join(bike_counters, by = "id") |>
  st_as_sf() |>
  ggplot() + 
  annotation_map_tile(type = "cartolight", zoomin = 0) + 
  geom_sf(aes(color = counts)) +
  theme_void()
```

spatially join bike counters and lanes
```{r}
bike_counts_2024 |> 
          summarize(.by = id, 
                    counts = sum(counts) / 365) |>
          left_join(bike_counters, by = "id") |>
  st_as_sf() |>
  st_join(bike_routes |>
              st_buffer(dist = 50),
          left = TRUE) |> 
  ggplot() + 
  annotation_map_tile(type = "cartolight",zoomin = 0) + 
  geom_sf(aes(color = facilitycl)) + 
  theme_void()
```

So counters are (mostly) near protected bike lanes


spatially join crashes and bike lanes?
First need to add a buffer around lanes
```{r}
buffers <- st_buffer(bike_routes, dist = 100)

collisions_routes <- st_join(collisions_2024, buffers) |>
  mutate(near_bike_route = if_else(!is.na(segmentid),1,0)) |>
  filter(near_bike_route == 1)
```

```{r}
ggplot() +
  annotation_map_tile(type = "cartolight", zoomin = 0) + 
  geom_sf(data = collisions_routes,
          aes(color = facilitycl),
          alpha = 0.5) +
  geom_sf(data = bike_routes,
          aes(color = facilitycl)) +
  theme_void() 
```



Collisions in my CD
```{r}
collisions_cd <- st_join(collisions_2024,nyc_cd) 

bike_routes_cd <- st_join(bike_routes,
                          nyc_cd |>
                            filter(boro_cd == 107),
                          left = FALSE)

collisions_cd |>
  filter(boro_cd == 107) |>
  ggplot() + 
  annotation_map_tile(type = "cartolight", zoomin = 0) + 
  geom_sf() + 
  geom_sf(data = bike_routes_cd,
          aes(color = facilitycl)) + 
  theme_void()
```


Density of bike lanes by borough
```{r}
bike_routes |>
  mutate(bike_lane_length = st_length(bike_routes)) |>
  group_by(boro) |>
  summarize(bike_lane_length = sum(bike_lane_length)) 

nyc |>
  mutate(area = st_area(nyc)) |>
  select(NAME, area)
```

## analysis ideas
- time period: 2024
- unit of analysis:
  - crash point. was crash in bike lane? what type of bike lane?
  - bike lane. how many crashes per mile of bike lane of different types? can we focus on high volume bike lanes?
  - neighborhood / CD. density of crashes and bike lanes per mile? as a fraction of total in the city?
  
I am worried about not being able to account for traffic volumes. of course there will be more crashes where there are more riders...Maybe focusing in on bike counters or a single neighborhood will help, but it is tricky.

# Main analysis

## 1. Crashes by bike lane type

Bike crashes by lane type (raw)

Categorizing each crash by whether it was within 100 feet of a bike lane
```{r}
buffers <- st_buffer(bike_routes, dist = 100)

collisions_routes <- st_join(collisions_2024, buffers,
                             largest = TRUE) |>
  mutate(near_bike_route = if_else(!is.na(segmentid),1,0)) 

collisions_routes |>
  group_by(facilitycl) |>
  summarize(n = n(),
            num_bike_injured = sum(number_of_cyclist_injured),
            num_bike_killed = sum(number_of_cyclist_killed)) |>
  st_drop_geometry() |>
  left_join(bike_routes |>
               mutate(length = as.numeric(st_length(geometry))) |>
               group_by(facilitycl) |>
               summarize(length_miles = sum(length)/5280) |>
               st_drop_geometry()
             ) |>
  mutate(crashes_per_mile = n / length_miles,
         injuries_per_mile = num_bike_injured / length_miles,
         deaths_per_mile = num_bike_killed / length_miles,
         killed_per_crash = num_bike_killed / n) 
  # writexl::write_xlsx("output/crash_route_table.xlsx")


```


## 2. By neighborhood 

crashes and bike lanes per area (and per population?)

May want to account for whether crashes were on bike lanes or not on ibke lanes!
```{r}
buffers <- st_buffer(bike_routes, dist = 100)

collisions_routes <- st_join(collisions_2024, buffers) |>
  mutate(near_bike_route = if_else(!is.na(segmentid),1,0)) 

cd_routes <- nyc_cd |>
  # start with community districts, get their areas
  filter(pop1 > 1000) |>
  mutate(area = as.numeric(st_area(geometry)) / (5280*5280)) |>
  # Spatially join to bike routes with length
  st_join(bike_routes |>
            mutate(bike_lane_length = as.numeric(st_length(bike_routes))/5280)) |>
  # aggregate by cd and bike lane type
  group_by(boro_cd, area, pop1) |>
  summarize(bike_lane_length = as.numeric(sum(bike_lane_length))) 


  # now join cd with collisions on routes
cd_collisions <- nyc_cd |>
  filter(pop1 > 1000) |>
  mutate(area = as.numeric(st_area(geometry)) / (5280*5280)) |>
  st_join(collisions_routes) |> 
  group_by(boro_cd, area, pop1) |>
  summarize(num_crashes = n(),
            crashes_on_routes = sum(near_bike_route)) 

cd_collisions_routes <- cd_routes |>
  st_drop_geometry() |>
  full_join(cd_collisions) |>
  st_as_sf() |>
  mutate(crashes_on_routes_per_mile_bike_lane = crashes_on_routes/bike_lane_length,
         crashes_per_mile_bike_lane = num_crashes / bike_lane_length,
         crashes_per_area = num_crashes / area,
         bike_lanes_per_area = bike_lane_length / area,
         crashes_per_pop = num_crashes / pop1,
         bike_lanes_per_pop = bike_lane_length / pop1) |>
  ungroup() |>
  mutate(crashes_pct = num_crashes / sum(num_crashes),
         bike_lanes_pct = bike_lane_length / sum(bike_lane_length),
         pop_pct = pop1 / sum(pop1))
```

Make some maps
```{r}
breaks <- classInt::classIntervals(cd_collisions_routes$crashes_per_area, n = 4, style = "quantile")$brks

map1 <- cd_collisions_routes |>
  ggplot() +
  geom_sf(aes(fill = crashes_per_area),
          alpha = 0.5) + 
  scale_fill_fermenter(name = NULL,
                       breaks = breaks,
                       labels = c("",round(breaks,1)[2:length(breaks)]),
                       type = "div",
                       palette = "RdYlBu",
                       guide = "colorbar") +
  labs(title ="A. Collisions per Square Mile") + 
  theme_void() + 
  theme(legend.position = "inside",
        legend.position.inside = c(0.3,0.7))

# move legend, add scale and north arrow
breaks <- classInt::classIntervals(cd_collisions_routes$bike_lanes_per_area, n = 4, style = "quantile")$brks

map2 <- cd_collisions_routes |>
  ggplot() +
  geom_sf(aes(fill = bike_lanes_per_area),
          alpha = 0.5) + 
  scale_fill_fermenter(name = NULL,
                       breaks = breaks,
                       labels = round(breaks,1),
                       type = "div",
                       palette = "RdYlBu",
                       guide = "colorbar") +
  labs(title = "B. Miles of Bike Lanes per Square Mile") + 
  theme_void() + 
  annotation_north_arrow(
    style = north_arrow_minimal,
    height = unit(.8, "cm"),
    width = unit(.6, "cm"),
    pad_x = unit(1.5, "in"), 
    pad_y = unit(0.2, "in"),
  ) + 
  annotation_scale(
    style = "ticks",
    pad_x = unit(2, "in"), 
    pad_y = unit(0.2, "in"),
    unit_category = "imperial") + 
  theme(legend.position = "inside",
        legend.position.inside = c(0.3,0.7))

map1 + map2 

ggsave("output/density_cd.jpg",
        width = 7, height = 5)
```

```{r}
breaks <- classInt::classIntervals(cd_collisions_routes$crashes_on_routes_per_mile_bike_lane, n = 5, style = "jenks")$brks

cd_collisions_routes |>
  ggplot() +
  geom_sf(aes(fill = crashes_on_routes_per_mile_bike_lane),
          alpha = 0.5) + 
  scale_fill_fermenter(name = NULL,
                       breaks = breaks,
                       labels = round(breaks,1),
                       type = "div",
                       palette = "RdYlBu",
                       guide = "colorbar") +
  labs(title = "Crashes on Bike Routes per Mile") + 
  theme_void() +
  annotation_north_arrow(
    style = north_arrow_minimal,
    height = unit(.8, "cm"),
    width = unit(.6, "cm"),
    pad_x = unit(2.5, "in"), 
    pad_y = unit(0.3, "in"),
  ) + 
  annotation_scale(
    style = "ticks",
    pad_x = unit(3, "in"), 
    pad_y = unit(0.3, "in"),
    unit_category = "imperial") + 
  theme(legend.position = "inside",
        legend.position.inside = c(0.3,0.7))

ggsave("output/crashes_per_mile.jpg",
       width = 6, height = 5)
```

bivariate map
```{r}
bi_cd <- bi_class(
  cd_collisions_routes,      # data
  x = bike_lanes_per_area,    # First variable: % below poverty line (will be classified into 3 quantile groups)
  y = crashes_per_area,    # Second variable: COVID death rate (also classified into 3 quantile groups)
  style = "quantile", # Method: quantile cuts; could also use "jenks"
  dim = 4              # Dimension: 3 x 3 classification (low/medium/high for each variable)
)

bi_map <- ggplot () +
  geom_sf(
    data = bi_cd, 
    aes(fill = bi_class), 
    color = NA, 
    linewidth = 0.1, 
    show.legend = FALSE
  ) +
   bi_scale_fill(
    pal = "DkBlue2", 
    dim = 4
  ) +
  geom_sf(
    data = nyc_cd,
    color = "black",
    fill = NA
  ) + theme_void() +
  labs(title = "C. Bike Lane and Crash Density") +
  theme(legend.position = "inside",
        legend.position.inside = c(0.3,0.7))

legend <- bi_legend(
  pal = "DkBlue2",
  dim = 4,
  xlab = "Bike Lane Density",
  ylab =  "Crash Density",
  size = 8
)  

# Add the bivariate legend to the map as an inset (using {patchwork})

map3 <- bi_map +
  inset_element(
    legend,              # the bivariate legend grob (created with bi_legend())
    left = 0.05,         # 5% from the left edge of the map
    bottom = 0.45,       # starting halfway up from the bottom
    right = 0.43,        # extends to 43% of map width
    top = 0.85           # extends to 85% of map height
  ) 


map1 + map2 + map3
ggsave("output/lane_crash_density_bivar.jpg",
       width = 12, height = 4)
```


Here, can see that as the bike lane density increases, the number of crashes increases.

This might indicate that the density of bike lanes and the density of crashes are both basically capturing the traffic level in a given area.

However, when we look at what percentage of crashes are on bike routes, the relationship is highly non-linear. When there are few bike lanes, the pct of crashes on bike lanes is low (as expected) and the percentage of crashes on bike langes increases ~linearly. However about 5 miles of bike lanes per square mile the percentage of crashes on routes levels off.
```{r}
p1 <- cd_collisions_routes |>
  ggplot(aes(x = bike_lanes_per_area, y = crashes_per_area)) +
  geom_point(color = "royalblue") +
  geom_smooth(method = "lm",
              color = "red") +
  labs(x = "Miles of Bike Lanes per Square Mile",
       y = "Collisions per Square Mile",
       subtitle = "Bike Collision and Lane Density by Community District") + 
  theme_minimal() 

cd_collisions_routes |>
  mutate(pct_on_routes = crashes_on_routes/num_crashes) |> 
  ggplot(aes(x = bike_lane_length, y = num_crashes)) +
  geom_point() +
  geom_smooth(method = "lm")

p2 <- cd_collisions_routes |>
  mutate(pct_on_routes = crashes_on_routes/num_crashes) |> 
  ggplot(aes(x = 10000*bike_lanes_per_pop, y = 10000*crashes_per_pop)) +
  geom_point(color = "royalblue") +
  geom_smooth(method = "lm",
              color = "red") +
  labs(x = "Miles of Bike Lanes per 10,000 people",
       y = "Collisions per 10,000 people",
       subtitle = "Bike Collision and Lanes per 10,000 Population") + 
  theme_minimal() 


cd_collisions_routes |>
  mutate(pct_on_routes = crashes_on_routes/num_crashes) |> 
  ggplot(aes(x = bike_lanes_per_area,y = 100 * pct_on_routes,)) +
  geom_point(color = "royalblue") +
  geom_smooth(color = "red",
              se = FALSE)+
  labs(x = "Miles of Bike Lanes per Square Mile",
       y = "Percent of Collisions on Bike Lanes",
       subtitle = "Collisions on Bike Lanes by Community District") + 
  ylim(0,100) + 
  theme_minimal() 

p1 + p2
ggsave("output/crash_per_area_cd.jpg",
       width = 8, height = 4)

ggsave("output/crash_per_area_scatter.jpg",
       p1,
       width = 6, height = 4)
```

I think the outline should be something like this
- crashes by bake lane type (table) DONE
- crahes per mile of bike lane at CD level (scatterplot)
- bike lanes per CD area
- crashes per CD area


## By neighborhood and bike lane type
```{r}
buffers <- st_buffer(bike_routes, dist = 100)

collisions_routes <- st_join(collisions_2024, buffers) |>
  mutate(near_bike_route = if_else(!is.na(segmentid),1,0)) 

cd_routes <- nyc_cd |>
  # start with community districts, get their areas
  filter(pop1 > 1000) |>
  mutate(area = as.numeric(st_area(geometry)) / (5280*5280)) |>
  # Spatially join to bike routes with length
  st_join(bike_routes |>
            mutate(bike_lane_length = as.numeric(st_length(bike_routes))/5280)) |>
  # aggregate by cd and bike lane type
  group_by(boro_cd, area, pop1, facilitycl) |>
  summarize(bike_lane_length = as.numeric(sum(bike_lane_length))) 


  # now join cd with collisions on routes
cd_collisions <- nyc_cd |>
  filter(pop1 > 1000) |>
  mutate(area = as.numeric(st_area(geometry)) / (5280*5280)) |>
  st_join(collisions_routes) |> 
  group_by(boro_cd, area, pop1, facilitycl) |>
  summarize(num_crashes = n(),
            crashes_on_routes = sum(near_bike_route)) 

cd_collisions_routes <- cd_routes |>
  st_drop_geometry() |>
  inner_join(cd_collisions) |>
  group_by(boro_cd) |>
  complete(geometry, facilitycl)|>
  st_as_sf() |>
  mutate(crashes_on_routes = if_else(is.na(crashes_on_routes),0,crashes_on_routes),
         crashes_on_routes_per_mile_bike_lane = crashes_on_routes/bike_lane_length,
         crashes_per_area = num_crashes / area,
         bike_lanes_per_area = bike_lane_length / area,
         crashes_per_lane_density = num_crashes / bike_lanes_per_area,
         crashes_per_pop = num_crashes / pop1,
         bike_lanes_per_pop = bike_lane_length / pop1) |>
  group_by(boro_cd) |>
  mutate(crashes_pct = num_crashes / sum(num_crashes,na.rm = T),
         bike_lanes_pct = bike_lane_length / sum(bike_lane_length,na.rm = T))
```

Some plots using this
```{r}
cd_collisions_routes |>
  group_by(boro_cd) |>
  mutate(total_crashes_per_area = sum(crashes_per_area,na.rm = T),
         total_crashes = sum(num_crashes,na.rm = T),
         total_lane_length = sum(bike_lane_length,na.rm = T),
         total_bike_lane_density= sum(bike_lanes_per_area,na.rm = T),
         total_crashes_per_mile_lane = total_crashes / total_lane_length,
         bike_lane_len_pct = bike_lane_length / sum(bike_lane_length,na.rm = T)) |>
  # filter(facilitycl == "Protected") |>
  ggplot(aes(x = bike_lane_len_pct, y = total_crashes_per_mile_lane)) +
  geom_point() +
  geom_smooth(method = "lm") + 
  facet_wrap(~facilitycl)

cd_collisions_routes |>
  filter(facilitycl != "Link") |>
  ggplot(aes(x = bike_lanes_per_area, y = crashes_per_area,
             color = facilitycl)) +
  geom_point() +
  geom_smooth(method = "lm")  +
  theme_minimal()

cd_collisions_routes |>
  filter(facilitycl != "Link") |>
  ggplot(aes(x = 100 * bike_lanes_pct, y = 100 * crashes_pct,
             color = facilitycl, size = bike_lane_length)) +
  geom_point(alpha = 0.5) +
  geom_smooth(size = 1,
              method = "lm",
              se = FALSE)  +
  scale_size_continuous(name = "Bike Lane Length (Miles)") + 
  scale_color_discrete(name = "Bike Lane Type") +
  labs(x = "Percent of Bike Lanes",
       y = "Percent of Crashes on Bike Lanes",
       title = "Percent of Crashes vs. Percent of Bike Lanes",
       subtitle = "By Community District") +
  theme_minimal()

ggsave("output/pct_crashes_lanes.jpg",
       width = 5,height = 3.5)

cd_collisions_routes |>
  group_by(boro_cd) |> 
  mutate(total_crashes_per_area = sum(crashes_per_area,na.rm = T),
         total_crashes = sum(num_crashes,na.rm = T),
         total_bike_lane_density= sum(bike_lanes_per_area,na.rm = T),
         bike_lane_len_pct = bike_lane_length / sum(bike_lane_length,na.rm = T),
         pct_on_routes = crashes_on_routes/sum(num_crashes)) |>
  # filter(facilitycl == "Protected") |>
  ggplot(aes(x = bike_lane_len_pct, y = total_crashes_per_area,
             size = total_bike_lane_density)) +
  geom_point() +
  facet_wrap(~facilitycl)

cd_collisions_routes |>
  ggplot(aes(x = factor(boro_cd),
             y = bike_lane_length,
             fill = facilitycl)) + 
  geom_bar(position = "fill", stat = "identity")

cd_collisions_routes |>
  group_by(boro_cd) |>
  mutate(bike_lane_len_pct = bike_lane_length / sum(bike_lane_length,na.rm = T)) |>
  ggplot(aes(x = bike_lane_len_pct, y = num_crashes)) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~ facilitycl)
```

Map of percentages
```{r}
centroids <- cd_collisions_routes |> 
  st_centroid() 

temp <- cd_collisions_routes |>
  group_by(boro_cd) |>
  summarize(bike_lane_length = sum(bike_lane_length,na.rm = T))

breaks <- classInt::classIntervals(temp$bike_lane_length, n = 5, style = "quantile")$brks

ggplot() + 
  geom_sf(data = cd_collisions_routes |>
            group_by(boro_cd) |>
            summarize(bike_lane_length = sum(bike_lane_length,na.rm = T)),
          aes(fill = bike_lane_length),
          alpha = 0.5) + 
  scale_fill_fermenter(name = "Total Miles of Bike lanes",
                       type = "div",
                       palette = "Spectral",
                       guide = "colorbar") +
  ggnewscale::new_scale_fill() +
  scatterpie::geom_scatterpie(
    data = cd_collisions_routes |>
      mutate(centroids = st_centroid(geometry),
             long = st_coordinates(centroids)[,1],
             lat = st_coordinates(centroids)[,2]) |>
      group_by(boro_cd) |>
      mutate(bike_lane_len_pct = bike_lane_length / sum(bike_lane_length,na.rm = T)) |>
      st_drop_geometry() |>
      group_by(boro_cd) |>
      complete(facilitycl) |>
      fill(long,lat, .direction = "downup") |>
      mutate(bike_lane_len_pct = if_else(is.na(bike_lane_len_pct),0,bike_lane_len_pct)) |>
      select(boro_cd, long, lat, facilitycl, bike_lane_len_pct) |>
      pivot_wider(names_from = facilitycl,
                  values_from = bike_lane_len_pct),
    aes(x = long, y = lat,group = boro_cd),
    cols = c("Protected","Conventional","Shared Lane","Link")
  ) +
  scale_fill_discrete(name = "Bike lane type") + 
  labs(title = "Distribution of Bike Lane types",
       subtitle = "By Total Mileage and Community District") + 
  theme_void() +
  annotation_north_arrow(
    style = north_arrow_minimal,
    height = unit(.8, "cm"),
    width = unit(.6, "cm"),
    pad_x = unit(2.5, "in"), 
    pad_y = unit(0.3, "in"),
  ) + 
  annotation_scale(
    style = "ticks",
    pad_x = unit(3, "in"), 
    pad_y = unit(0.3, "in"),
    unit_category = "imperial") + 
  theme(legend.position = "inside",
        legend.position.inside = c(0.3,0.7))

ggsave("output/lane_dist.jpg",
       width = 5, height = 6)
```

## 3 panel map
4 panel map? or maybe 3 dropping "link"
```{r}
t1 <- cd_collisions_routes |> 
  filter(facilitycl == "Protected")

breaks <- classInt::classIntervals(t1$crashes_on_routes_per_mile_bike_lane, 
                                   n = 5, style = "jenks")$brks

m1 <- t1 |>
  ggplot() +
  geom_sf(aes(fill = crashes_on_routes_per_mile_bike_lane)) + 
  geom_sf_text(data = \(d) d |>
                  arrange(-crashes_on_routes_per_mile_bike_lane) |>
                  head(3),
                aes(label = boro_cd),
                color = "black") + 
  scale_fill_fermenter(name = "Crashes per Mile",
                       breaks = breaks,
                       labels = round(breaks,1),
                       type = "div",
                       palette = "RdYlBu",
                       guide = "colorbar") +
  labs(title = "Bike Crashes on Route per Mile by Bike Route Type",
       subtitle = "Protected") + 
  theme_void() +
  theme(legend.position = "inside",
        legend.position.inside = c(0.3,0.7))

t2 <- cd_collisions_routes |> 
  filter(facilitycl == "Conventional")

breaks <- classInt::classIntervals(t2$crashes_on_routes_per_mile_bike_lane, 
                                   n = 5, style = "jenks")$brks

m2 <- t2 |>
  ggplot() +
  geom_sf(aes(fill = crashes_on_routes_per_mile_bike_lane)) + 
  geom_sf_text(data = \(d) d |>
                  arrange(-crashes_on_routes_per_mile_bike_lane) |>
                  head(3),
                aes(label = boro_cd),
                color = "black") + 
  scale_fill_fermenter(name = "Crashes per Mile",
                       breaks = breaks,
                       labels = round(breaks,1),
                       type = "div",
                       palette = "RdYlBu",
                       guide = "colorbar") +
  labs(subtitle = "Conventional") + 
  theme_void()+
  theme(legend.position = "inside",
        legend.position.inside = c(0.3,0.7))

t3 <- cd_collisions_routes |> 
  filter(facilitycl == "Shared Lane")

breaks <- classInt::classIntervals(t3$crashes_on_routes_per_mile_bike_lane, 
                                   n = 5, style = "jenks")$brks

m3 <- t3 |>
  ggplot() +
  geom_sf(aes(fill = crashes_on_routes_per_mile_bike_lane)) + 
  geom_sf_text(data = \(d) d |>
                  arrange(-crashes_on_routes_per_mile_bike_lane) |>
                  head(3),
                aes(label = boro_cd),
                color = "black") + 
  scale_fill_fermenter(name = "Crashes per Mile",
                       breaks = breaks,
                       labels = round(breaks,1),
                       type = "div",
                       palette = "RdYlBu",
                       guide = "colorbar") +
  labs(subtitle = "Shared Lane") + 
  theme_void() + 
  annotation_north_arrow(
    style = north_arrow_minimal,
    height = unit(.8, "cm"),
    width = unit(.6, "cm"),
    pad_x = unit(1.5, "in"), 
    pad_y = unit(0.2, "in"),
  ) + 
  annotation_scale(
    style = "ticks",
    pad_x = unit(2, "in"), 
    pad_y = unit(0.2, "in"),
    unit_category = "imperial") + 
  theme(legend.position = "inside",
        legend.position.inside = c(0.3,0.7))

m1 + m2 + m3

ggsave("output/crashes_per_mile_route.jpg",
       width = 10, height = 6)
```

bivariate map x3
```{r}
bi_cd_lane <- cd_collisions_routes |>
  filter(facilitycl %in% c("Protected","Conventional","Shared Lane")) |>
  group_by(facilitycl) |>
  group_modify(
    ~bi_class(.x,
              x = bike_lanes_per_area,    # First variable: % below poverty line (will be classified into 3 quantile groups)
              y = crashes_per_area,    # Second variable: COVID death rate (also classified into 3 quantile groups)
              style = "quantile", # Method: quantile cuts; could also use "jenks"
              dim = 3              # Dimension: 3 x 3 classification (low/medium/high for each variable)
    )) |>
  ungroup() |>
  st_as_sf()

bi_map <- ggplot () +
  geom_sf(
    data = bi_cd_lane, 
    aes(fill = bi_class), 
    color = NA, 
    linewidth = 0.1, 
    show.legend = FALSE
  ) +
   bi_scale_fill(
    pal = "GrPink", 
    dim = 3
  ) +
  geom_sf(
    data = nyc_cd,
    color = "black",
    fill = NA
  ) + 
  theme_void() +
  facet_wrap(~facilitycl,
             ncol = 2) + 
  labs(title = "Bike Lane and Crash Density by Lane Type") + 
  theme(strip.text = element_text(hjust = 0.5,
                                  size = rel(1.2)))

legend <- bi_legend(
  pal = "GrPink",
  dim = 3,
  xlab = "Bike Lane Density",
  ylab =  "Crash Density",
  size = 10
)  

# Add the bivariate legend to the map as an inset (using {patchwork})

bi_map +
  inset_element(
    legend,              # the bivariate legend grob (created with bi_legend())
    left = 0.6,     
    right = 0.9,     
    bottom = 0.1,      
    top = 0.4         
  )

ggsave("output/lanetype_crash_bivar.jpg",
       width = 6, height = 6)
```


## KDE

```{r}
kde <- collisions_2024 |> 
  hotspot_kde(
    cell_size        = 2500,   # Diameter (in map units, e.g., meters) of each hex cell; larger value → coarser grid
    grid_type        = "hex", # Use a hexagonal grid for density calculation (could also use "rect")
    bandwidth_adjust = 0.5    # Scale factor for the kernel bandwidth; <1 tightens (more local), >1 smooths more broadly
  ) |> 
  st_intersection(nyc_cd |>
                    filter(pop1 > 1000)) |>  # Clip the KDE hex grid to Chicago’s boundary, dropping cells outside the city
  mutate(
    kde = round(kde, 1)        # Round the density estimate to one decimal place for cleaner display
  )

jenks_breaks <- classInt::classIntervals(kde$kde, n = 5, style = "jenks")$brks


ggplot() +
  geom_sf(
    data = nyc_cd,
    fill = NA,
    color = "black",
    linewidth = 0.2,
  ) + 
  geom_sf(
    data = kde,
    aes(fill = kde),
    col = NA) +
  scale_fill_fermenter(
    type = "div",
    palette = "RdYlBu",
    breaks = jenks_breaks,
    labels = round(jenks_breaks,1),
    guide = "colorbar"
  ) +
  theme_void()
```




