---
title: "Bike Collision GIS analysis"
author: "Ravi Brenner"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(sf)
library(ggspatial)
library(sfhotspot)
```

Importing and cleaning data

Bike routes
```{r}
bike_routes <- st_read("data/nyc_bike_routes.shp")  |>
  st_transform(crs = 2263) |>
  filter(year(instdate) <= 2024,
         status == "Current") |> # only want bike routes that existed in 2024 or earlier
  mutate(ft_facilit = case_match(ft_facilit,
                                 "<Null>" ~ NA,
                                 "NA" ~ NA,
                                 .default = ft_facilit),
         tf_facilit = case_match(tf_facilit,
                                 "<Null>" ~ NA,
                                 "NA" ~ NA,
                                 .default = tf_facilit),
         boro = case_match(boro,
                           1 ~ "Manhattan",
                           2 ~ "Bronx",
                           3 ~ "Brooklyn",
                           4 ~ "Queens",
                           5 ~ "Staten Island"),
         facilitycl = case_match(facilitycl,
                                 "I" ~ "Protected",
                                 "II" ~ "Conventional",
                                 "III" ~ "Shared Lane",
                                 "L" ~ "Link"),
         facilitycl = factor(facilitycl,
                             levels = c("Protected",
                                        "Conventional",
                                        "Shared Lane",
                                        "Link")))
```

Collisions for 2024
```{r}
collisions_2024 <- read_csv("data/collisions_2024.csv",
                            na = c("NA")) |>
  drop_na(latitude,longitude) |>
  filter(latitude != 0,
         longitude != 0,
         number_of_cyclist_injured > 0 | number_of_cyclist_killed > 0) |>
  st_as_sf(coords = c("longitude","latitude"),
           crs = 4326) |>
  st_transform(crs = st_crs(bike_routes))
```

Bike counts and counters
```{r}
bike_counts_2024 <- read_csv("data/bike_counts_2024.csv")

bike_counters <- read_csv("data/bike_counters.csv") |>
  filter(latitude != 0,
         longitude != 0) |>
  st_as_sf(coords = c("longitude","latitude"),
           crs = 4326) |>
  st_transform(crs = st_crs(bike_routes))
```

NYC boundaries
```{r}
nyc <- st_read("data/Counties_Shoreline.shp") |>
  filter(NAME %in% c("Bronx","Richmond","Kings","Queens","New York")) |>
  st_transform(crs = st_crs(bike_routes))
```

Community districts
```{r}
nyc_cd <- st_read("data/nyc_cd.shp") |>
  st_transform(crs = st_crs(bike_routes))
```



# Preliminary map

```{r}
ggplot() + 
  annotation_map_tile(type = "cartolight", zoomin = 0) +
  geom_sf(data = collisions_2024,
          color = "lightblue",
          alpha = 0.5) +
  geom_sf(data = bike_routes,
          aes(color = facilitycl)) +
  scale_color_discrete(name = "Bike Lane Type") + 
  labs(title = "Bike Lanes and Collisions Involving a Bicycle in New York City, 2024",
       caption = "Author: Ravi Brenner\nSource: NYC Open Data\nNote: crash locations shown in blue may represent multiple crashes at that location") + 
  theme_void()  +
  theme( 
    plot.caption.position = "plot",
    plot.caption = element_text(hjust = 1,
                                vjust = 20),
    legend.position = "inside",
    legend.position.inside = c(0.2,0.7),
    legend.key.height = unit(0.6, "cm"), 
    legend.text = element_text(size = 10),
    legend.background = element_rect(
      fill = alpha("white", 0.8),
      color = NA)
  ) + 
  annotation_north_arrow(
    height = unit(.8, "cm"),
    width = unit(.6, "cm"),
    pad_x = unit(2.5, "in"), 
    pad_y = unit(0.35, "in"),
    style = north_arrow_minimal
  ) +
  annotation_scale(
    style = "ticks",
    pad_x = unit(1.5, "in"),
    pad_y = unit(0.4, "in"),
    unit_category = "imperial",
    width_hint = 0.1
  )

ggsave(
  filename = "output/prelim_map_1.jpg",
  width = 7,
  height = 7
  )
```

join collisions with city neighborhoods
```{r}
collisions_cd <- st_join(nyc_cd, collisions_2024)

temp <- collisions_cd |>
  group_by(boro_cd) |>
  summarize(num_crashes = n()) |>
  mutate(area = as.numeric(st_area(geometry) / (5280*5280)),
         crashes_per_area = num_crashes / area) 

jenks_breaks <- classInt::classIntervals(temp$crashes_per_area, style = "jenks", n = 4)$brks

temp |>
  ggplot() + 
  annotation_map_tile(type = "cartolight", zoomin = 0) + 
  geom_sf(aes(fill = crashes_per_area),
          alpha = 0.5) + 
  scale_fill_viridis_c(name = "Crashes per Square Mile",
                       breaks = jenks_breaks,
                       labels = c(0.15,14.5,37.4,58.1,123.5)) +
  labs(title = "Bike Crashes by Community District, New York City, 2024",
       caption = "Author: Ravi Brenner\nSource: NYC Open Data") + 
  theme_void() +
    theme( 
    plot.caption.position = "plot",
    plot.caption = element_text(hjust = 1,
                                vjust = 15),
    legend.position = "inside",
    legend.position.inside = c(0.2,0.7),
    legend.key.height = unit(0.6, "cm"), 
    legend.text = element_text(size = 10),
    legend.background = element_rect(
      fill = alpha("white", 0.5),
      color = NA),
    legend.margin = margin(b = 5)
  ) + 
  annotation_north_arrow(
    height = unit(.8, "cm"),
    width = unit(.6, "cm"),
    pad_x = unit(2.5, "in"), 
    pad_y = unit(0.35, "in"),
    style = north_arrow_minimal
  ) +
  annotation_scale(
    style = "ticks",
    pad_x = unit(1.5, "in"),
    pad_y = unit(0.4, "in"),
    unit_category = "imperial",
    width_hint = 0.1
  )

ggsave(
  filename = "output/prelim_map_2.jpg",
  width = 7,
  height = 7
  )
```

# EDA maps
locations of bike counters
```{r}
ggplot() +
  annotation_map_tile(type = "cartolight", zoomin = 0) + 
  geom_sf(data = bike_counters) + 
  geom_sf(data = bike_routes,
          aes(color = facilitycl)) +
  theme_void()
```

Bike counters with counts
```{r}
bike_counts_2024 |> 
  summarize(.by = id, 
            counts = sum(counts) / 365) |>
  left_join(bike_counters, by = "id") |>
  st_as_sf() |>
  ggplot() + 
  annotation_map_tile(type = "cartolight", zoomin = 0) + 
  geom_sf(aes(color = counts)) +
  theme_void()
```

spatially join bike counters and lanes
```{r}
bike_counts_2024 |> 
          summarize(.by = id, 
                    counts = sum(counts) / 365) |>
          left_join(bike_counters, by = "id") |>
  st_as_sf() |>
  st_join(bike_routes |>
              st_buffer(dist = 50),
          left = TRUE) |> 
  ggplot() + 
  annotation_map_tile(type = "cartolight",zoomin = 0) + 
  geom_sf(aes(color = facilitycl)) + 
  theme_void()
```

So counters are (mostly) near protected bike lanes


spatially join crashes and bike lanes?
First need to add a buffer around lanes
```{r}
buffers <- st_buffer(bike_routes, dist = 100)

collisions_routes <- st_join(collisions_2024, buffers) |>
  mutate(near_bike_route = if_else(!is.na(segmentid),1,0)) |>
  filter(near_bike_route == 1)
```

```{r}
ggplot() +
  annotation_map_tile(type = "cartolight", zoomin = 0) + 
  geom_sf(data = collisions_routes,
          aes(color = facilitycl),
          alpha = 0.5) +
  geom_sf(data = bike_routes,
          aes(color = facilitycl)) +
  theme_void() 
```



Collisions in my CD
```{r}
collisions_cd <- st_join(collisions_2024,nyc_cd) 

bike_routes_cd <- st_join(bike_routes,
                          nyc_cd |>
                            filter(boro_cd == 106),
                          left = FALSE)

collisions_cd |>
  filter(boro_cd == 106) |>
  ggplot() + 
  annotation_map_tile(type = "cartolight", zoomin = 0) + 
  geom_sf() + 
  geom_sf(data = bike_routes_cd,
          aes(color = facilitycl)) + 
  theme_void()
```


Density of bike lanes by neighborhood
```{r}
bike_routes |>
  mutate(bike_lane_length = st_length(bike_routes)) |>
  group_by(boro) |>
  summarize(bike_lane_length = sum(bike_lane_length)) 

nyc |>
  mutate(area = st_area(nyc)) |>
  select(NAME, area)
```

## analysis ideas
- time period: 2024
- unit of analysis:
  - crash point. was crash in bike lane? what type of bike lane?
  - bike lane. how many crashes per mile of bike lane of different types? can we focus on high volume bike lanes?
  - neighborhood / CD. density of crashes and bike lanes per mile? as a fraction of total in the city?
  
I am worried about not being able to account for traffic volumes. of course there will be more crashes where there are more riders...Maybe focusing in on bike counters or a single neighborhood will help, but it is tricky.

