---
title: "Bike Collision GIS analysis"
author: "Ravi Brenner"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(sf)
library(ggspatial)
```

Importing and cleaning data

Bike routes
```{r}
bike_routes <- st_read("data/nyc_bike_routes.shp")  |>
  st_transform(crs = 2263) |>
  filter(year(instdate) <= 2024,
         status == "Current") |> # only want bike routes that existed in 2024 or earlier
  mutate(ft_facilit = case_match(ft_facilit,
                                 "<Null>" ~ NA,
                                 "NA" ~ NA,
                                 .default = ft_facilit),
         tf_facilit = case_match(tf_facilit,
                                 "<Null>" ~ NA,
                                 "NA" ~ NA,
                                 .default = tf_facilit),
         boro = case_match(boro,
                           1 ~ "Manhattan",
                           2 ~ "Bronx",
                           3 ~ "Brooklyn",
                           4 ~ "Queens",
                           5 ~ "Staten Island"),
         facilitycl = case_match(facilitycl,
                                 "I" ~ "Protected",
                                 "II" ~ "Conventional",
                                 "III" ~ "Shared Lane",
                                 "L" ~ "Link"),
         facilitycl = factor(facilitycl,
                             levels = c("Protected",
                                        "Conventional",
                                        "Shared Lane",
                                        "Link")))
```

Collisions for 2024
```{r}
collisions_2024 <- read_csv("data/collisions_2024.csv",
                            na = c("NA")) |>
  drop_na(latitude,longitude) |>
  filter(latitude != 0,
         longitude != 0,
         number_of_cyclist_injured > 0 | number_of_cyclist_killed > 0) |>
  st_as_sf(coords = c("longitude","latitude"),
           crs = 4326) |>
  st_transform(crs = st_crs(bike_routes))
```

Bike counts and counters
```{r}
bike_counts_2024 <- read_csv("data/bike_counts_2024.csv")

bike_counters <- read_csv("data/bike_counters.csv") |>
  filter(latitude != 0,
         longitude != 0) |>
  st_as_sf(coords = c("longitude","latitude"),
           crs = 4326) |>
  st_transform(crs = st_crs(bike_routes))
```

NYC boundaries
```{r}
nyc <- st_read("data/Counties_Shoreline.shp") |>
  filter(NAME %in% c("Bronx","Richmond","Kings","Queens","New York")) |>
  st_transform(crs = st_crs(bike_routes))
```

Community districts
```{r}
# populations
cd_pop <- readxl::read_excel("data/2020CensusData/nyc_decennialcensusdata_2010_2020_change-core-geographies_v2.xlsx",
                             sheet = "2020") |>
  janitor::clean_names() |> 
  filter(geo_type == "CD") |>
  mutate(geo_id = as.numeric(geo_id)) |>
  select(geo_id, pop1)

nyc_cd <- st_read("data/nyc_cd.shp") |>
  st_transform(crs = st_crs(bike_routes)) |>
  inner_join(cd_pop, 
             by = join_by(boro_cd == geo_id))
```



# Preliminary map

```{r}
ggplot() + 
  annotation_map_tile(type = "cartolight", zoomin = 0) +
  geom_sf(data = collisions_2024,
          color = "lightblue",
          alpha = 0.5) +
  geom_sf(data = bike_routes,
          aes(color = facilitycl)) +
  scale_color_discrete(name = "Bike Lane Type") + 
  labs(title = "Bike Lanes and Collisions Involving a Bicycle in New York City, 2024",
       caption = "Author: Ravi Brenner\nSource: NYC Open Data\nNote: crash locations shown in blue may represent multiple crashes at that location") + 
  theme_void()  +
  theme( 
    plot.caption.position = "plot",
    plot.caption = element_text(hjust = 1,
                                vjust = 20),
    legend.position = "inside",
    legend.position.inside = c(0.2,0.7),
    legend.key.height = unit(0.6, "cm"), 
    legend.text = element_text(size = 10),
    legend.background = element_rect(
      fill = alpha("white", 0.8),
      color = NA)
  ) + 
  annotation_north_arrow(
    height = unit(.8, "cm"),
    width = unit(.6, "cm"),
    pad_x = unit(2.5, "in"), 
    pad_y = unit(0.35, "in"),
    style = north_arrow_minimal
  ) +
  annotation_scale(
    style = "ticks",
    pad_x = unit(1.5, "in"),
    pad_y = unit(0.4, "in"),
    unit_category = "imperial",
    width_hint = 0.1
  )

ggsave(
  filename = "output/prelim_map_1.jpg",
  width = 7,
  height = 7
  )
```

join collisions with city neighborhoods
```{r}
collisions_cd <- st_join(nyc_cd, collisions_2024)

temp <- collisions_cd |>
  group_by(boro_cd) |>
  summarize(num_crashes = n()) |>
  mutate(area = as.numeric(st_area(geometry) / (5280*5280)),
         crashes_per_area = num_crashes / area) 

jenks_breaks <- classInt::classIntervals(temp$crashes_per_area, style = "jenks", n = 4)$brks

temp |>
  ggplot() + 
  annotation_map_tile(type = "cartolight", zoomin = 0) + 
  geom_sf(aes(fill = crashes_per_area),
          alpha = 0.5) + 
  scale_fill_viridis_c(name = "Crashes per Square Mile",
                       breaks = jenks_breaks,
                       labels = c(0.15,14.5,37.4,58.1,123.5)) +
  labs(title = "Bike Crashes by Community District, New York City, 2024",
       caption = "Author: Ravi Brenner\nSource: NYC Open Data") + 
  theme_void() +
    theme( 
    plot.caption.position = "plot",
    plot.caption = element_text(hjust = 1,
                                vjust = 15),
    legend.position = "inside",
    legend.position.inside = c(0.2,0.7),
    legend.key.height = unit(0.6, "cm"), 
    legend.text = element_text(size = 10),
    legend.background = element_rect(
      fill = alpha("white", 0.5),
      color = NA),
    legend.margin = margin(b = 5)
  ) + 
  annotation_north_arrow(
    height = unit(.8, "cm"),
    width = unit(.6, "cm"),
    pad_x = unit(2.5, "in"), 
    pad_y = unit(0.35, "in"),
    style = north_arrow_minimal
  ) +
  annotation_scale(
    style = "ticks",
    pad_x = unit(1.5, "in"),
    pad_y = unit(0.4, "in"),
    unit_category = "imperial",
    width_hint = 0.1
  )

ggsave(
  filename = "output/prelim_map_2.jpg",
  width = 7,
  height = 7
  )
```

# EDA maps
locations of bike counters
```{r}
ggplot() +
  annotation_map_tile(type = "cartolight", zoomin = 0) + 
  geom_sf(data = bike_counters) + 
  geom_sf(data = bike_routes,
          aes(color = facilitycl)) +
  theme_void()
```

Bike counters with counts
```{r}
bike_counts_2024 |> 
  summarize(.by = id, 
            counts = sum(counts) / 365) |>
  left_join(bike_counters, by = "id") |>
  st_as_sf() |>
  ggplot() + 
  annotation_map_tile(type = "cartolight", zoomin = 0) + 
  geom_sf(aes(color = counts)) +
  theme_void()
```

spatially join bike counters and lanes
```{r}
bike_counts_2024 |> 
          summarize(.by = id, 
                    counts = sum(counts) / 365) |>
          left_join(bike_counters, by = "id") |>
  st_as_sf() |>
  st_join(bike_routes |>
              st_buffer(dist = 50),
          left = TRUE) |> 
  ggplot() + 
  annotation_map_tile(type = "cartolight",zoomin = 0) + 
  geom_sf(aes(color = facilitycl)) + 
  theme_void()
```

So counters are (mostly) near protected bike lanes


spatially join crashes and bike lanes?
First need to add a buffer around lanes
```{r}
buffers <- st_buffer(bike_routes, dist = 100)

collisions_routes <- st_join(collisions_2024, buffers) |>
  mutate(near_bike_route = if_else(!is.na(segmentid),1,0)) |>
  filter(near_bike_route == 1)
```

```{r}
ggplot() +
  annotation_map_tile(type = "cartolight", zoomin = 0) + 
  geom_sf(data = collisions_routes,
          aes(color = facilitycl),
          alpha = 0.5) +
  geom_sf(data = bike_routes,
          aes(color = facilitycl)) +
  theme_void() 
```



Collisions in my CD
```{r}
collisions_cd <- st_join(collisions_2024,nyc_cd) 

bike_routes_cd <- st_join(bike_routes,
                          nyc_cd |>
                            filter(boro_cd == 107),
                          left = FALSE)

collisions_cd |>
  filter(boro_cd == 107) |>
  ggplot() + 
  annotation_map_tile(type = "cartolight", zoomin = 0) + 
  geom_sf() + 
  geom_sf(data = bike_routes_cd,
          aes(color = facilitycl)) + 
  theme_void()
```


Density of bike lanes by borough
```{r}
bike_routes |>
  mutate(bike_lane_length = st_length(bike_routes)) |>
  group_by(boro) |>
  summarize(bike_lane_length = sum(bike_lane_length)) 

nyc |>
  mutate(area = st_area(nyc)) |>
  select(NAME, area)
```

## analysis ideas
- time period: 2024
- unit of analysis:
  - crash point. was crash in bike lane? what type of bike lane?
  - bike lane. how many crashes per mile of bike lane of different types? can we focus on high volume bike lanes?
  - neighborhood / CD. density of crashes and bike lanes per mile? as a fraction of total in the city?
  
I am worried about not being able to account for traffic volumes. of course there will be more crashes where there are more riders...Maybe focusing in on bike counters or a single neighborhood will help, but it is tricky.

# Main analysis

## Crashes by bike lane type

Bike crashes by lane type (raw)

Categorizing each crash by whether it was within 100 feet of a bike lane
```{r}
buffers <- st_buffer(bike_routes, dist = 100)

collisions_routes <- st_join(collisions_2024, buffers,
                             largest = TRUE) |>
  mutate(near_bike_route = if_else(!is.na(segmentid),1,0)) 

collisions_routes |>
  group_by(facilitycl) |>
  summarize(n = n(),
            num_bike_injured = sum(number_of_cyclist_injured),
            num_bike_killed = sum(number_of_cyclist_killed)) |>
  st_drop_geometry() |>
  left_join(bike_routes |>
               mutate(length = as.numeric(st_length(geometry))) |>
               group_by(facilitycl) |>
               summarize(length_miles = sum(length)/5280) |>
               st_drop_geometry()
             ) |>
  mutate(crashes_per_mile = n / length_miles,
         injuries_per_mile = num_bike_injured / length_miles,
         deaths_per_mile = num_bike_killed / length_miles,
         killed_per_crash = num_bike_killed / n) 
```

## crashes by bike lane - counters

focus on lanes near counters

Having some issues with this--not going to use it in final analysis
```{r}
# bike counts at each counter
bike_counts <- bike_counts_2024 |> 
  summarize(.by = id, 
            counts = sum(counts)) |>
  left_join(bike_counters, by = "id") |>
  group_by(counter) |>
  mutate(keep_row = case_when(
    n() == 1 ~ 1,
    n() > 1 & str_detect(name,"Bike") ~ 1,
    n() > 1 & str_detect(name,"Bicycle") ~ 1,
    .default =0
  )) |> 
  filter(keep_row == 1) |>
  st_as_sf() 

# find bike lane closest to each counter
buffers <- st_buffer(bike_routes,100)

counters_routes <- st_join(bike_counts,buffers,
                             largest = FALSE) |> 
  filter(!is.na(counter))

# Find collisions closest to bike lanes
buffers <- st_buffer(bike_routes, dist = 100)

collisions_routes <- st_join(collisions_2024, buffers,
                             largest = TRUE) |>
  mutate(near_bike_route = if_else(!is.na(segmentid),1,0)) |>
  st_drop_geometry() |>
  group_by(bikeid) |>
  summarize(num_crashes = n(),
            num_bike_injured = sum(number_of_cyclist_injured),
            num_bike_killed = sum(number_of_cyclist_killed))


counter_crashes <- counters_routes |>
  st_drop_geometry() |>
  group_by(id) |>
  mutate(max_count = max(counts),
            n = n(),
            avg_count = max_count / n) |>
  inner_join(collisions_routes,
             by = "bikeid") |> 
  mutate(crashes_per_count = n / avg_count) |>
  inner_join(bike_routes) |> 
  st_as_sf() |>
  group_by(bikeid) |>
  mutate(bike_lane_length = as.numeric(sum(st_length(geometry)))/5280,
         count_miles = avg_count * bike_lane_length,
         crashes_per_count_mile = n /count_miles) |>
  filter(count_miles > 10000) 

counter_crashes |>
  group_by(bikeid) |>
  summarize(name = list(unique(str_c(street,"_",fromstreet,"_",tostreet))),
            facilitycl = unique(facilitycl),
            crashes = max(num_crashes),
            rider_volume = max(avg_count),
            bike_lane_length = max(bike_lane_length),
            crashes_per_count_mile = max(crashes_per_count_mile)) |>
  arrange(-crashes_per_count_mile)

counter_crashes |>
  ggplot() +
  annotation_map_tile(type = "cartolight", zoomin = 0) +
  geom_sf(aes(color = 10000*crashes_per_count_mile)) +
  scale_color_viridis_c(name = "Crashes per 10,000 annual rider-miles") +
  theme_void()

```

## By neighborhood 

crashes and bike lanes per area (and per population?)

May want to account for whether crashes were on bike lanes or not on ibke lanes!
```{r}
buffers <- st_buffer(bike_routes, dist = 100)

collisions_routes <- st_join(collisions_2024, buffers) |>
  mutate(near_bike_route = if_else(!is.na(segmentid),1,0)) 

cd_routes <- nyc_cd |>
  # start with community districts, get their areas
  filter(pop1 > 1000) |>
  mutate(area = as.numeric(st_area(geometry)) / (5280*5280)) |>
  # Spatially join to bike routes with length
  st_join(bike_routes |>
            mutate(bike_lane_length = as.numeric(st_length(bike_routes))/5280)) |>
  # aggregate by cd and bike lane type
  group_by(boro_cd, area, pop1) |>
  summarize(bike_lane_length = as.numeric(sum(bike_lane_length))) 


  # now join cd with collisions on routes
cd_collisions <- nyc_cd |>
  filter(pop1 > 1000) |>
  mutate(area = as.numeric(st_area(geometry)) / (5280*5280)) |>
  st_join(collisions_routes) |> 
  group_by(boro_cd, area, pop1) |>
  summarize(num_crashes = n(),
            crashes_on_routes = sum(near_bike_route)) 

cd_collisions_routes <- cd_routes |>
  st_drop_geometry() |>
  full_join(cd_collisions) |>
  st_as_sf() |>
  mutate(crashes_on_routes_per_mile_bike_lane = crashes_on_routes/bike_lane_length,
         crashes_per_area = num_crashes / area,
         bike_lanes_per_area = bike_lane_length / area,
         crashes_per_pop = num_crashes / pop1,
         bike_lanes_per_pop = bike_lane_length / pop1) |>
  ungroup() |>
  mutate(crashes_pct = num_crashes / sum(num_crashes),
         bike_lanes_pct = bike_lane_length / sum(bike_lane_length),
         pop_pct = pop1 / sum(pop1))
```

Make some maps
```{r}
breaks <- classInt::classIntervals(cd_collisions_routes$crashes_per_area, n = 5, style = "jenks")$brks

cd_collisions_routes |>
  ggplot() +
  geom_sf(aes(fill = crashes_per_area),
          alpha = 0.5) + 
  scale_fill_viridis_c(name = NULL,
                       breaks = breaks,
                       labels = round(breaks,1)) +
  labs(title = "Bike Collisions per Square Mile",
       subtitle = "By NYC Community Districts, 2024") + 
  theme_void()

# move legend, add scale and north arrow
breaks <- classInt::classIntervals(cd_collisions_routes$bike_lanes_per_area, n = 5, style = "jenks")$brks

cd_collisions_routes |>
  ggplot() +
  geom_sf(aes(fill = bike_lanes_per_area),
          alpha = 0.5) + 
  scale_fill_viridis_c(name = NULL,
                       breaks = breaks,
                       labels = round(breaks,1)) +
  labs(title = "Bike Lane Density by Community District, 2024",
       subtitle = "Miles of Bike Lanes per Square Mile") + 
  theme_void()
```


Here, can see that as the bike lane density increases, the number of crashes increases.

This might indicate that the density of bike lanes and the density of crashes are both basically capturing the traffic level in a given area.

However, when we look at what percentage of crashes are on bike routes, the relationship is highly non-linear. When there are few bike lanes, the pct of crashes on bike lanes is low (as expected) and the percentage of crashes on bike langes increases ~linearly. However about 5 miles of bike lanes per square mile the percentage of crashes on routes levels off.
```{r}
cd_collisions_routes |>
  ggplot(aes(x = bike_lanes_per_area, y = crashes_per_area)) +
  geom_point() +
  geom_smooth(method = "lm")

cd_collisions_routes |>
  mutate(pct_on_routes = crashes_on_routes/num_crashes) |> 
  ggplot(aes(x = bike_lane_length, y = num_crashes)) +
  geom_point() +
  geom_smooth(method = "lm")

cd_collisions_routes |>
  mutate(pct_on_routes = crashes_on_routes/num_crashes) |> 
  ggplot(aes(x = bike_lanes_per_area,y = pct_on_routes)) +
  geom_point() +
  geom_smooth()
```

I think the outline should be something like this
- crashes by bake lane type (table) DONE
- crahes per mile of bike lane at CD level (scatterplot)
- bike lanes per CD area
- crashes per CD area


## By neighborhood and bike lane type
```{r}
buffers <- st_buffer(bike_routes, dist = 100)

collisions_routes <- st_join(collisions_2024, buffers) |>
  mutate(near_bike_route = if_else(!is.na(segmentid),1,0)) 

cd_routes <- nyc_cd |>
  # start with community districts, get their areas
  filter(pop1 > 1000) |>
  mutate(area = as.numeric(st_area(geometry)) / (5280*5280)) |>
  # Spatially join to bike routes with length
  st_join(bike_routes |>
            mutate(bike_lane_length = as.numeric(st_length(bike_routes))/5280)) |>
  # aggregate by cd and bike lane type
  group_by(boro_cd, area, pop1, facilitycl) |>
  summarize(bike_lane_length = as.numeric(sum(bike_lane_length))) 


  # now join cd with collisions on routes
cd_collisions <- nyc_cd |>
  filter(pop1 > 1000) |>
  mutate(area = as.numeric(st_area(geometry)) / (5280*5280)) |>
  st_join(collisions_routes) |> 
  group_by(boro_cd, area, pop1, facilitycl) |>
  summarize(num_crashes = n(),
            crashes_on_routes = sum(near_bike_route)) 

cd_collisions_routes <- cd_routes |>
  st_drop_geometry() |>
  full_join(cd_collisions) |>
  st_as_sf() |>
  mutate(crashes_on_routes_per_mile_bike_lane = crashes_on_routes/bike_lane_length,
         crashes_per_area = num_crashes / area,
         bike_lanes_per_area = bike_lane_length / area,
         crashes_per_pop = num_crashes / pop1,
         bike_lanes_per_pop = bike_lane_length / pop1) |>
  ungroup() |>
  mutate(crashes_pct = num_crashes / sum(num_crashes),
         bike_lanes_pct = bike_lane_length / sum(bike_lane_length),
         pop_pct = pop1 / sum(pop1))
```

Some plots using this
```{r}
cd_collisions_routes |>
  ggplot(aes(x = bike_lanes_per_area, y = crashes_per_area,
             group = facilitycl, color = facilitycl)) +
  geom_point() +
  geom_smooth(method = "lm")

cd_collisions_routes |>
  mutate(pct_on_routes = crashes_on_routes/num_crashes) |> 
  ggplot(aes(x = bike_lane_length, y = num_crashes,
             group = facilitycl, color = facilitycl)) +
  geom_point() +
  geom_smooth()

cd_collisions_routes |>
  ggplot(aes(x = factor(boro_cd),
             y = bike_lane_length,
             fill = facilitycl)) + 
  geom_bar(position = "fill", stat = "identity")

cd_collisions_routes |>
  group_by(boro_cd) |>
  mutate(bike_lane_len_pct = bike_lane_length / sum(bike_lane_length,na.rm = T)) |>view()
  ggplot(aes(x = bike_lane_len_pct, y = num_crashes)) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~ facilitycl)
```

```{r}
cd_collisions_routes |>
  group_by(boro_cd) |>
  mutate(bike_lane_len_pct = bike_lane_length / sum(bike_lane_length,na.rm = T)) |>
  ggplot() + 
  geom_sf(aes(fill = bike_lane_len_pct),
          alpha = 0.5) +
  scale_fill_viridis_c()+
  theme_void() +
  facet_wrap( ~ facilitycl)
```


